The role of standing variation in geographic convergent adaptation

### 2014-09-29 standing-variation-paper.tex
hopefully, close to final changes

@@

\paragraph{The confusing signal of geographic convergent evolution.}
As we have argued in \citet{ralphcoop2010} the ease with which
geographic convergent adaptation occurs means that we should
incorporate it more widely into our thinking about the genetic basis
of adaptation. For example, the absence of European skin pigmentation alleles in
ancient DNA from European recovered from several thousand years ago
has led to the suggestion that these individuals had dark
pigmentation \citep{olalde2014derived,lazaridis2013ancient, Wilde2014}. 
However, given our results and the partially convergent
basis of skin pigmentation between Europeans and East Asians
\citep{norton-human-pigmentation,edwards2010pigmentation} it seems just as plausible that these ancient individuals
adapted to high latitudes via a different complement of ``light-skin'' pigmentation alleles;
we have no evidence either way.
Such convergence may considerably complicate the 
exploration of phenotypes and adaptation among populations using variants mapped in a
limited set of populations \citep{Berg:14}.

@@

\paragraph{Are species held together by widespread selective sweeps?}
Our results touch on an old debate on the evolutionary coherence of species. 
Mayr and many others have argued that species are coherent
evolutionary units because they are united by shared gene flow \citep[pages 521--522 in][]{Mayr:SpeciesEvol}. 
However, this argument has been questioned by a number of authors 
based on relatively high levels of differentiation, and low rates of dispersal, in many species \citep{EhrlichRaven:69,Levin:79}.
Even if gene flow is not high enough to prevent neutral differentiation or local adaptation, 
a number of authors have argued that species are cohesive 
if gene flow is high enough for globally selected alleles 
(and their hitchhiking haplotypes)
to spread across entire species 
\citep[see also][]{Rieseberg2001,MorjanRieseberg:04,Ellstrand2014}.
At present, large scale genotyping and sequencing projects, 
along with more sophisticated methods, 
are highlighting ever more signals of gene flow between populations and species \citep{Patterson:12,SousaHey:13,Hellenthal:14}.
However, our work on geographic convergent adaptation
\citep[see also][]{ralphcoop2010,RalphCoop:14} 
suggests that species should often adapt to
wide-spread selection pressures through convergent evolution rather
than waiting for a single allele to migrate across the range.
% over length-scales determined by our characteristic length.

@@

Such ideas are not completely new and represent a perhaps logical consequence of
an allopatric, or at least parapatric, view of the geography of speciation.
However, it is worth revisiting this idea as geographically broad population genomic
sampling allows us to return to themes
in biogeography. 
Along similar lines, Futuyma has argued that much of the adaptive differentiation within species,
e.g.\ adaptation to local conditions, 
may be ephemeral and subject to loss due to local extinction and the mixing 
following the collapse of population structure \citep{Futuyma:10,FUTUYMA:87}. 
Futuyma offered this as an explanation of the pattern of punctuated equilibrium \citep{eldredgegould72}, 
and argued that the observation of stasis and rapid anagenesis associated with speciation 
were consistent with micro-evolution. 
Futuyma argued that despite rapid adaptation over short time-scales, we
may observe morphological stasis in the fossil record as much of this
adaptation is lost to local extinction and the collapse of population structure
\citep[see also][]{lieberman:96,Eldredge:05}. 
Furthermore, he suggests that speciation
may act as ratchet to prevent the loss of differentiation, acting to
maintain adaptive changes among populations, and prevent their loss by interbreeding. 
At face value the rate of species formation seems too low to contribute to this process. 
However, \citep{Rosenblum:12}, and many others, 
have argued that the rate of speciation may well be quite high,
but that the majority of incipient species do not persist long due to reabsorption or extinction. 
Repeated bouts of extinction and speciation will send waves of alleles
to fixation along particular lineages. 
%Ranges change sufficiently rapidly that
%range size could change substantially over the sojourn time of most alleles 
%that will eventually become fixed in the species, making it likely that.

@@

Finally, we close by noting that range expansion and speciation 
are obviously not separate from adaptive differentiation. 
The invasion of new geographic areas may lead to a burst of adaptive differentiation, 
at least in a subset of genes, and speciation may be associated with rapidly adaption to novel environments. 
Conversely, if the geographic spread of adaptive alleles within ranges is slow, 
e.g.\ if they only offer a local advantage or if they are selective excluded, 
may allow Dobzhansky-Muller incompatibilities to arise within species, 
effectively offering a mechanism for hybrid incompatibilities to evolve in parapatry,
and fracturing the range \citep{Bank:12,Kondrashov:03,bierne2011coupling}. 
The alleles that act as components in many of the Dobzhansky-Muller incompatibilities studied to date 
are geographically restricted \citep[see][]{Cutter:12}. 
Therefore, it seems possible that populations within species may be tending towards speciation, 
and that as outlined here that this may drive some proportion of molecular divergence.

\subsubsection*{Acknowledgements}
We thank Michael Turelli, Jon Seger and the Coop lab for helpful conversations.

>>>>

Note that the conflicting roles of $\rho$ and $\sigma$ mean thar even in
species where levels of neutral differentiation are low, geographic
convergent adaptation may be common. This is because low levels of
neutral genetic differentiation can be due to high population
densities rather than high dispersal distances, and high population
densities would allow convergent adaptation. As such geographic convergent
evolution may be common even in species with little neutral population structure.

@@

\paragraph{The confusing signal of geographic convergent evolution.}
As we have argued in \citet{ralphcoop2010} the ease with which
geographic convergent adaptation occurs means that we should
incorporate it more widely into our thinking about the genetic basis
of adaptation. For example, the absence of European skin pigmentation alleles in
ancient DNA from European recovered from several thousand years ago
has led to the suggestion that these individuals had dark
pigmentation \citep{olalde2014derived,lazaridis2013ancient, Wilde2014}. 
However, given our results and the partially convergent
basis of skin pigmentation between Europeans and East Asians
\citep{norton-human-pigmentation,edwards2010pigmentation} it seems just as plausible that these ancient individuals
adapted to high latitudes via a different complement of ``light-skin'' pigmentation alleles;
to our knowledge, we have no strong evidence either way.
Such convergence may considerably complicate the 
exploration of phenotypes and adaptation among populations using variants mapped in a
limited set of populations \citep{Berg:14}.

@@

\paragraph{Are species held together by widespread selective sweeps?}
Our results touch on an old debate on the evolutionary coherence of species. 
Mayr and many others have argued that species are coherent
evolutionary units because they are united by shared gene flow \citep[pages 521--522 in][]{Mayr:SpeciesEvol}. 
However, this argument has been questioned by a number of authors 
based on relatively high levels of differentiation, and low rates of dispersal, in many species \citep{EhrlichRaven:69,Levin:79}.
Even if gene flow is not high enough to prevent neutral differentiation or local adaptation, 
a number of authors have argued that species are cohesive 
if gene flow is high enough for globally selected alleles 
(and their hitchhiking haplotypes)
to spread across entire species 
\citep[see also][]{Rieseberg2001,MorjanRieseberg:04,Ellstrand2014}.
At present, large scale genotyping and sequencing projects, 
along with more sophisticated methods, 
are highlighting ever more signals of gene flow between populations and species \citep{Patterson:12,SousaHey:13,Hellenthal:14}.
However, our work on geographic convergent adaptation
\citep[see also][]{ralphcoop2010,RalphCoop:14} 
suggests that species should often adapt to
wide-spread selection pressures through convergent evolution rather
than waiting for a single allele to migrate across the range.

% over length-scales determined by our characteristic length.

@@

Such ideas are not completely new and represent a perhaps logical consequence of
an allopatric or parapatric view of the biogeography of speciation.
However, it is worth revisiting this idea as geographically broad population genomic
sampling allows us to return to themes
in biogeography. 
Along similar lines, Futuyma has argued that much of the adaptive differentiation within species,
e.g.\ adaptation to local conditions, 
may be ephemeral and subject to loss due to local extinction and the mixing 
following the collapse of population structure \citep{Futuyma:10,FUTUYMA:87}. 
Futuyma offered this as an explanation of the pattern of punctuated equilibrium \citep{eldredgegould72}, 
and argued that the observation of stasis and rapid anagenesis associated with speciation 
were consistent with micro-evolution. 
Futuyma argued that despite rapid adaptation over short time-scales, we
may observe morphological stasis in the fossil record as much of this
adaptation is lost to local extinction and the collapse of population structure
\citep[see also][]{lieberman:96,Eldredge:05, Futuyma:10}. 
Furthermore, he suggests that speciation
may act as ratchet to prevent the loss of differentiation, acting to
maintain adaptive changes among populations, and prevent their loss by interbreeding. 
At face value the rate of species formation seems too low to contribute to this process. 
However, \citep{Rosenblum:12}, and many others, 
have argued that the rate of speciation may well be quite high,
but that the majority of incipient species do not persist long due to
reabsorption or extinction. 
Changes in range size, due to local extinction, can also be very rapid 
on the time-scales over which alleles may spread on the landscape \citep{gaston2003structure,HEWITT1996}.
Repeated bouts of extinction and speciation will send waves of alleles
to fixation along particular lineages. 
%Ranges change sufficiently rapidly that
%range size could change substantially over the sojourn time of most alleles 
%that will eventually become fixed in the species, making it likely that.

@@

Finally, we close by noting that range expansion and speciation 
are obviously not separate from adaptive differentiation. 
The invasion of new geographic areas may lead to a burst of adaptive differentiation, 
at least in a subset of genes, and speciation may be associated with rapidly adaption to novel environments. 
Conversely, if the geographic spread of adaptive alleles within ranges is slow, 
e.g.\ if they only offer a local advantage or if they are selective excluded, 
may allow Dobzhansky-Muller incompatibilities to arise within species, 
effectively offering a mechanism for hybrid incompatibilities to evolve in parapatry,
and fracturing the range \citep{Bank:12,Kondrashov:03,bierne2011coupling}. 
The alleles that act as components in many of the Dobzhansky-Muller incompatibilities studied to date 
are geographically restricted \citep[see][]{Cutter:12}. 
Therefore, it seems possible that populations within species may often be tending towards speciation, 
and that as outlined here that this may drive some proportion of molecular divergence.

\subsubsection*{Acknowledgements}
We thank Michael Turelli, Jon Seger, and the Coop lab for helpful conversations.
